Estimating epidemiological quantities from repeated cross-sectional prevalence measurements

### 2022-02-24 docs/paper.Rmd
Draft writeup (#28)
* first draft of writeup

* add authors

* example fixes

* fix figure

* move to docs

* prettier table and move figures

* fix typo and update table

* first draft of writeup

* add authors

* example fixes

* fix figure

* move to docs

* prettier table and move figures

* fix typo

* abstract

* fixes

* add interpretation of biological parameters

* some more references and minor changes

We used Gaussian Process (GP) priors to ensure smoothness of the estimates and deal with data gaps, whereby alternatively either $I(t)$ is has a GP prior with exponential quadratic kernel.
$$
\begin{aligned}
I(t) &\sim \text{logit}(i_0 + i(t))\\
i(t) &\sim \text{GP}(t)
\end{aligned}
$$
where $i_0$ is the estimated mean of the GP, or the GP prior is applied to higher order differences, for example the growth rate such as
$$
i(t) - i(t - 1) \sim \text{GP}(t)
$$
which implies that growth, rather than incidence, remains at an estimated mean level in the absence of data, usually leading to better real-time performance [@abbott2020].
The results shown in this paper were obtained using this formulation with a GP prior on the growth rate.

@@

When additionally using antibodies we convolve the modelled infections $I(t)$ as well as input data on vaccinations $Y_{\text{V}, t}$ with distributions quantifying the delay to generating detectable antibodies following infection (by default set to 4 weeks for both infection and vaccination), yielding potentially antibody-generating time series from infection $I^{\text{A}}$ and $V^{\text{A}}$.
We then calculate antibodies from infection as
$$
A^{\text{I}}(t) =
A^{\text{I}}(t - 1) +
\beta I^{\text{A}}(t)  (1 - A(t - 1))^k -
\gamma_\text{I} A^{\text{I}}(t - 1)
$$
and antibodies from vaccination as
$$
A^{\text{V}}(t) =
A^{\text{V}}(t - 1) +
\delta V^{\text{A}}(t) (1 - A(t - 1))^l -
 \gamma_\text{V} A^{\text{V}}(t - 1)
$$
with the total population proportion with antibodies given as the sum of the two,
$$
A(t) = A^{\text{I}}(t)  + A^{\text{V}}(t)
$$

@@

The model was implemented in _stan_ and using the _cmdstanr_ R package.
All code needed to reproduce the results shown in this paper is available at https://github.com/epiforecasts/inc2prev.

@@

```{r params-table}
biol_params %>%  kable()
  ## kable(caption = "Estimates and 90% credible intervals (CIs, as quantiles of the posterior distribution) of biological parameters.")
  ##       booktabs = TRUE) %>%
  ## kable_styling()
```

Posterior estimates of recovered biological parameters are shown in Table \@ref{tab:params-table}.
Some of the parameter estimates show high levels of correlation suggesting issues of identifiability (Figure \@ref(fig:pairs-plot)).

@@

Our estimated parameters of antibody dynamics are averages across a various combinations of vaccine types and individual factors that are known to affect immunological responses to either infection or vaccination, particularly age [@ward2022]. 
We estimated that `r 100 - signif(params[name == "beta"]$median, 2) * 100`% (90% credible interval, CI: `r 100 - signif(params[name == "beta"]$q95, 2) * 100`--`r 100 - signif(params[name == "beta"]$q5, 2) * 100`) of individuals do not seroconvert after infection, compared to 24% estimated from the same study population, but also lower estimates such as 10% in a different study [@gudbjartsson2020]. 
We further estimated that `r signif(params[name == "delta"]$median, 2) * 100`% (90% CI: `r signif(params[name == "delta"]$q95, 2) * 100`--`r signif(params[name == "delta"]$q5, 2) * 100`) of individuals seroconverted following vaccination, in line with high such proportions estimated in healthcare workers [@eyre2021].
Our estimates of waning correspond to an average duration of immunity of `r signif(1 / params[name == "gamma (infection)"]$median, 2)` days (90% CI: `r signif(params[name == "gamma (infection)"]$q95, 2)`--`r signif(params[name == "gamma (infection)"]$q5, 2) `) following infection and `r signif(1 / params[name == "gamma (vaccination)"]$median, 2)` days (90% CI: `r signif(params[name == "gamma (vaccination)"]$q95, 2)`--`r signif(params[name == "gamma (vaccination)"]$q5, 2)`) following vaccination.

@@

There is enormous potential for understanding epidemiological dynamics from repeated cross-sectional surveys.
Where the generation interval distribution is the same or close to the distribution of detectability after infection, this could be done using recently developed methods for unified modelling of incidence and prevalence [@pakkanen2021].
The methods presented here and related ones could be applied to other infections monitored in a similar way, and thus in combination with such data collection and publication become a tool for monitoring epidemic and endemic infectious diseases in the future.

@@

We thank Thomas House for insightful comments on this work, and the Office for National Statistics for making the data sets publicly available.

>>>>

We used Gaussian Process (GP) priors to ensure smoothness of the estimates and deal with data gaps, whereby alternatively either $I(t)$ is has a GP prior with exponential quadratic kernel.
$$
\begin{aligned}
I(t) &\sim \text{logit}(i_0 + i(t))\\
i(t) &\sim \text{GP}(t)
\end{aligned}
$$
where $i_0$ is the estimated mean of the GP, or the GP prior is applied to higher order differences, for example the growth rate such as
$$
i(t) - i(t - 1) \sim \text{GP}(t)
$$
which implies that growth, rather than incidence, remains at an estimated mean level in the absence of data, usually leading to better real-time performance [@abbott2020].
The results shown here were obtained using this formulation with a GP prior on the growth rate.

@@

When additionally using antibodies we convolve the modelled infections $I(t)$ as well as input data on vaccinations $Y_{\text{V}, t}$ with distributions quantifying the delay to generating detectable antibodies following infection (by default set to 4 weeks for both infection and vaccination), yielding potentially antibody-generating time series from infection $I_{\text{A}}$ and $V_{\text{A}}$.
We then calculate antibodies from infection as
$$
A_{\text{I}}(t) =
A_{\text{I}}(t - 1) +
\beta I_{\text{A}}(t)  (1 - A(t - 1))^k -
\gamma_\text{I} A_{\text{I}}(t - 1)
$$
and antibodies from vaccination as
$$
A_{\text{V}}(t) =
A_{\text{V}}(t - 1) +
\delta V_{\text{A}}(t) (1 - A(t - 1))^l -
 \gamma_\text{V} A_{\text{V}}(t - 1)
$$
with the total population proportion with antibodies given as the sum of the two,
$$
A(t) = A_{\text{I}}(t)  + A_{\text{V}}(t)
$$

@@

The model was implemented in _Stan_ and using the _cmdstanr_ R package [@gabry2021;@standev2022].
All code needed to reproduce the results shown here is available at https://github.com/epiforecasts/inc2prev.

@@

```{r params-table}
biol_params %>% 
  kbl(caption = paste0(
    "Estimates and credible intervals (CIs, as quantiles of the posterior",
    "distribution) of biological parameters."
    ), booktabs = TRUE) %>%
  kable_styling()
```

Posterior estimates of recovered biological parameters are shown in Table \@ref(tab:params-table).
Some of the parameter estimates show high levels of correlation suggesting issues of identifiability (Figure \@ref(fig:pairs-plot)).

@@

Our estimated parameters of antibody dynamics are averages across a various combinations of vaccine types and individual factors that are known to affect immunological responses to either infection or vaccination, particularly age [@ward2022]. 
We estimated that `r 100 - signif(params[name == "beta"]$median, 2) * 100`% (90% credible interval, CI: `r 100 - signif(params[name == "beta"]$q95, 2) * 100`--`r 100 - signif(params[name == "beta"]$q5, 2) * 100`) of individuals do not seroconvert after infection, compared to 24% estimated from the same study population, but also lower estimates such as 10% in a different study [@gudbjartsson2020]. 
We further estimated that `r signif(params[name == "delta"]$median, 2) * 100`% (90% CI: `r signif(params[name == "delta"]$q95, 2) * 100`--`r signif(params[name == "delta"]$q5, 2) * 100`) of individuals seroconverted following vaccination, in line with high such proportions estimated in healthcare workers [@eyre2021].
Our estimates of waning correspond to an average duration of immunity of `r signif(1 / params[name == "gamma (infection)"]$median, 2)` days (90% CI: `r signif(1 / params[name == "gamma (infection)"]$q95, 2)`--`r signif(1 / params[name == "gamma (infection)"]$q5, 2) `) following infection and `r signif(1 / params[name == "gamma (vaccination)"]$median, 2)` days (90% CI: `r signif(1 / params[name == "gamma (vaccination)"]$q95, 2)`--`r signif(1 / params[name == "gamma (vaccination)"]$q5, 2)`) following vaccination.

@@

There is enormous potential for understanding epidemiological dynamics from repeated cross-sectional surveys, whether to identify current or past infection [@metcalf2016].
Where the generation interval distribution is the same or close to the distribution of detectability after infection, this could be done using recently developed methods for unified modelling of incidence and prevalence [@pakkanen2021].
The methods presented here and related ones could be applied to other infections monitored in a similar way, and thus in combination with such data collection and publication become a tool for monitoring epidemic and endemic infectious diseases in the future.

@@

We thank Thomas House for insightful comments on this work, and the Office for National Statistics for making the data sets publicly available. We acknowledge funding by the Wellcome Trust (210758/Z/18/Z).
